---
title: "MotorTrend Coursera Project"
author: "Julie Foster"
date: "October 23, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Project Instructions

 You work for Motor Trend, a magazine about the automobile industry. Looking at a 
data set of a collection of cars, they are interested in exploring the 
 relationship between a set of variables and miles per gallon (MPG) (outcome). 
 They are particularly interested in the following two questions:     
    *	“Is an automatic or manual transmission better for MPG”    
    *	"Quantify the MPG difference between automatic and manual transmissions"
    

## Executive Summary

 To answer these questions, we will do the following:     
 1. Perform an exploratory data analysis     
 2. Fit multivariate models and determine best model selection  
 3. Examine the residuals and perform diagnostics     
 4. Summarize findings     

From this analysis it is shown that cars with manual transmission was better for the MPG term than those with automatic transmission.  According to the model we selected, using a manual transmission had an advantage of approximately 3 miles per gallon over the automatic transmission.


## 1. Exploratory Data Analysis
 We begin by loading the data and exploring the details:
 
```{r}
data(mtcars)
head(mtcars)
```

To look at some of the data, we will plot the two main variables of interest: "mpg" and the style of transmission "am" (0 = automatic, 1 = manual):

```{r}
boxplot(mpg ~ am, data = mtcars, col= "steelblue", 
        xlab = "Transmission (0 = automatic, 1 = manual)", 
        ylab = " Miles Per Gallon", 
        main = "Boxplot of MPG vs. Transmission")
```

##2. Model Selection
To begin the model selection process, we will use a stepwise regression.

Starting with a regression of all the variables:

```{r}
mtfit <- lm(mpg ~., mtcars)
summary(mtfit)$coef
```

From this we can see a strong variation in p-values.  Selecting the five lowest values, we will begin by fitting the next model with "disp", "hp", "wt", "qsec", and "am":

```{r}
mtfit5 <- lm(mpg ~ factor(am) + disp + hp + wt + qsec, mtcars)
summary(mtfit5)$coef
summary(mtfit5)$adj.r.squared
```

Here we can see that the Adjusted R-squared value is 0.8375, which means that almost 84% of the variance between the variables can be explained by the model.  The p-values, however, show that the coefficients of "disp" and "hp"" are not significant.

Selecting the three lowest p-values from mtfit5, the next model uses: 

```{r}
mtfit3 <- lm(mpg ~ factor(am) + wt + qsec, mtcars)
summary(mtfit3)$coef
summary(mtfit5)$adj.r.squared
```

From here, the Adjusted R-Squared value is 0.8336,
 which means the model can explain 83% of the variance of the
mpg variable and all of the p-values are significant.

And finally, using a simple linear model, fitting just the "am" and the "mpg":

```{r}
mtfit1 <- lm(mpg ~ factor(am), mtcars)
summary(mtfit1)$coef
summary(mtfit1)$adj.r.squared
```

This shows that on average, a car has a mean of 17.147 mpg with an automatic transmission, and (17.147 + 7.245 = ) 24.392 with a manual transmission.

The Adjusted R-squared value is 0.3385, which means that the 
model can explain about 34% of the variance of the MPG variable. This model has the lowest R-squared value and should be rejected as a valid model for this data.

From the three models we used: mtfit5, mtfit3, and mtfit1, we selected mtfit3 as the best model because of the Adjusted R-squared value of 0.8336 and the significant p-values of the 
three regressors.

##3. Examine Residuals
The next step involves using a plot to visualize the residuals. We have to make sure there is no pattern in residuals, which would indicate that there may be unaccounted variables in our model:

```{r}
plot(mtfit3, which=1)
```

The points are pretty well randomly scattered with no discernable pattern and show that our fit is robust.

Finally, we need to confirm that the residuals have a normal distribution. We do this by examining the QQ plot:

```{r}
plot (mtfit3, which=2)
```

It can be seen that the residuals mimic a normal 
distribution and therefore our statistical inferences are 
valid.

##4. Summary
After running an exploratory analysis on the data, we selected the best model, mtfit3, and examined the residuals. The effect of using a manual transmission is better for the "mpg" term. The interpretation of the estimated coefficients from mtfit3 show that when holding the weight("wt") and the 1/4 mile time("qsec") constant, the average change of approximately 3 miles per gallon for the using a manual transmission over an automatic transmission.
